---
title: "Three variable data and hierarchical model "
output: pdf_document
author: "Hugo Maruri-Aguilar"
date: "08/04/2021"
---

This document has a small example to illustrate the lasso methodology and the `R` code I previously prepared. I run first standard lasso from library `lars` and then with in-house
programs replicate the analysis. Then I run the hierarchical 
versions (H, W, S) of lasso analysis. An Appendix describes the main functions used.

# Data preparation

This is the initial synthetic data set, stored in variable `DAT`. The column headers
indicate the exponent term, i.e. `010` is $x_2$. Note that data columns are already centered around the mean. This centering of columns and response will remove the need to consider intercept in the rest of the analysis.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
source("C:/maruriwork/documents/bostonmanor/code/codigofuente.R")
setwd("C:/maruriwork/documents/bostonmanor/examples/")
DAT<-scan(file="smalldata.txt"); dim(DAT)<-c(4,7)
DAT0<-DAT<-t(DAT)
LD<-diag(rep(1,3));
colnames(DAT)<-c(nombres(LM=LD),"y");
```

```{r}
DAT 
```

Now add columns with interactions $x_1x_2$ and $x_1x_3$, which are also centered around their means. The matrix `DATX` has the regressors, while `DATY` has the response values. Column names of `DATX` are updated.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
DATY<-matrix(ncol=1,DAT[,4]); DATX<-DAT[,-4]
DATX<-cbind(DATX,DAT[,1]*DAT[,2],DAT[,1]*DAT[,3]); 
DATX<-scale(x=DATX,center=TRUE,scale=FALSE)
## the model
LD<-diag(rep(1,3)); LD<-rbind(LD,c(1,1,0),c(1,0,1))
colnames(DATX)<-nombres(LM=LD); colnames(DATY)<-"y"
```

```{r}
cbind(DATX,DATY)
```

This is the model, to be read as exponents of term in each row. Note absence of intercept.
```{r}
LD
```

\pagebreak

# Initial LASSO analysis

## Standard lasso analysis with function `lars` from the same library. 

```{r,echo=!FALSE,warning=FALSE,message=FALSE}
library(lars)
A<-lars(x=DATX,y=DATY,type="lasso",normalize=FALSE,intercept=FALSE)
```

The rows in the table below correspond to breakpoints determined 
by values of $\lambda$ at which there is a change 
in the piecewise linear trajectories of coefficients. 
The first column in the table are 
these breakpoint values of $\lambda$; then the column entries are lasso
estimates $\beta^L(\lambda)$, that is coefficients of model terms
` `r colnames(DATX)` `; the last column is the criterion
$L=||Y-X\beta||_2^2+\lambda||\beta||_1$. I did some post processing 
of code outputs `A$beta` and `A$lambda` before showing the table.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
BBA<-cbind(c(A$lambda,0), (as.matrix(A$beta)) ,
           A$RSS/2+c(A$lambda,0)*apply(abs(A$beta),1,sum)  )
BBA<-BBA[nrow(BBA):1,]; colnames(BBA)<-nombracoeffsmatrix(LM = LD)
round(BBA,4)
```

In standard Lasso like above, the maximum value of $\lambda$ for the
analysis is determined by $\max_i |X^TY|_i$. This is the value at
which all the coefficients shrink to zero. For
these data, this is the maximum of {`r abs(t(DATX)%*%DATY)`}. 
The breakpoints of $\lambda$ above are determined automatically by `lars` and note that the analysis 
does not consider the hierarchical structure of the model terms.

##  Lasso with `quadprog`, single orthant

In preparation for later analyses, here is the least squares fit $\hat\beta$, corresponding to the coefficients in the first row in the table above. We also compute the vector of signs of $\hat\beta$, to be used in quadrant optimization.
```{r,echo=!FALSE}
bols<-solve(t(DATX)%*%DATX)%*%t(DATX)%*%DATY; t(bols) ## LSE estimate
sign(bols)->c0; t(c0) ## the quadrant of the LSE estimate
```

```{r,echo=FALSE,eval=FALSE}
## lm(DATY~DATX-1)$coefficients  ## Same OLS quantity, not evaluated
c(bols*32) ## for nice printing
```

The lasso using the quadrant function. The output of the function is 
the lasso estimate $\hat\beta^L(\lambda)$ and the achieved value $L$. 
Here we do two computations of the table above for $\lambda=0$ and
for $\lambda=`r A$lambda[length(A$lambda)]`$.

```{r}
minconoineq(CM=diag(c(c0)),XM=DATX,YM=DATY,lamm=0)
minconoineq(CM=diag(c(c0)),XM=DATX,YM=DATY,lamm=A$lambda[length(A$lambda)])
```

\pagebreak

## Lasso with `quadprog`, list of $\lambda$ and neighboring orthants

Here we use the values of $\lambda$={`r sort(round(c(A$lambda,0),4))`} from the `lars` previous analysis. From the initial orthant (`r c0`), the function
below computes neighbors by switching one orthant at a time:

```{r}
vecinos(c(c0))
```

In below, there are $|LM|+1$ optimizations for every value of $\lambda$. This is the size of the entry given in `GG`. Note that the absence of the parameter `AA`
means the optimization is performed without constraints apart from the orthants.

```{r,eval=!FALSE}
round(lassocono(XM=DATX, YM=DATY, GG=vecinos(c(c0)), 
                 lambdaval=sort(c(A$lambda,0))  )[,-c(2:6)],5)
```


## Lasso with `quadprog`, list of $\lambda$ and all orthants

Using the same function, the search below uses all orthants. This is achieved by not specifying orthants `GG` and thus for every value of $\lambda$
there are $2^{|LM|}$ optimizations. As above, the parameter `AA` was not used.

```{r,eval=!FALSE}
round(lassocono(XM=DATX, YM=DATY, 
                 lambdaval=sort(c(A$lambda,0))  )[,-c(2:6)],5)
```

Note that without specifying constraints matrix `AA`, both approaches give 
exactly the
same results.


\pagebreak

## Timings for both calls 

This initial example is by no means an expensive computation. Nevertheless, as there are 
`r length(A$lambda)+1` values of $\lambda$, the first call (neighbor)
does `r (length(A$lambda)+1)*(ncol(DATX)+1)` optimizations which is compared
against `r (length(A$lambda)+1)*(2^ncol(DATX))` optimizations of the
second call (full orthant). Thus for this example, the full orthant search is
`r round((2^ncol(DATX))/(ncol(DATX)+1),3)`
times more expensive than neighbor orthant search.


```{r}
library(tictoc)
tic("Neighboring orthants")
B0<-lassocono(XM=DATX, YM=DATY, GG=vecinos(c(c0)), lambdaval=sort(c(A$lambda,0)))
toc()
tic("All orthants")
B0<-lassocono(XM=DATX, YM=DATY, lambdaval=sort(c(A$lambda,0)))
toc()
```

\newpage

# Hierarchical lasso (H)

## The Hasse diagram

The function `hassediagram` retrieves the dependences between terms. 
Each row corresponds to model term; the columns are indexed by 
variables and. In a given row, a nonzero entry indexes which term is 
divisible by the corresponding row, and the position in the column is the
remainder. 
In other words, every non zero value corresponds to an edge
in the diagram, and the edge is drawn between the term indexed by the 
entry and the term inding the row.

```{r}
HD<-hassediagram(LM=LD); row.names(HD)<-nombres(LM=LD); HD
```

For the model of this example, the Hasse diagram has `r nrow(LD)` nodes (number of rows) and `r length(c(HD)[c(HD)!=0])` edges (number of non-zero entries).
Of the nodes, `r sum(apply(X=HD!=0,MARGIN=1,FUN=sum)==2)` node (`r rownames(HD)[apply(X=HD!=0,MARGIN=1,FUN=sum)==2]`) has two descendant
terms ; other `r sum(apply(X=HD!=0,MARGIN=1,FUN=sum)==1)` nodes 
(`r rownames(HD)[apply(X=HD!=0,MARGIN=1,FUN=sum)==1]`)
have only one descendant each and
`r sum(apply(X=HD==0,MARGIN=1,FUN=prod))` nodes 
(`r rownames(HD)[apply(X=HD==0,MARGIN=1,FUN=prod)==1]`)
have no descendant nodes. 

## Matrix of constraints for hierarchy $H$

We now build the analysis using the $H$ set of constraints, stemming
directly from the edges of the Hasse diagram. The function `restrict` 
gives the 
*S* type of hierarchy using option `type=1`; 
the type *W* using option `type=2` and 
the type *H* using option `type=3`; it 
calls internally `hassediagram` so no need to compute the diagram
beforehand. 

```{r}
AA<-restrict(LM=LD,type=3,weight=FALSE); colnames(AA)<-nombres(LM = LD); AA
```


### Constrained least squares orthant

The first approach uses only the orthant from least squares (`r c0`).

```{r}
bcc<-minconoineq(CM=diag(c(c0)),XM = DATX, YM = DATY,AA = AA,lamm = 0)
Ball<-matrix(nrow=1,c(0,bcc)); colnames(Ball)<-nombracoeffsmatrix(LM = LD)
round(Ball,5); 
```

### Constrained least squares neighbors of ols orthant 

We use the less expensive search over neighboring orthants of the ordinary least squares (`r c0`). 

```{r}
matrix(nrow=1,lassocono(XM=DATX, YM=DATY,AA=AA,GG=vecinos(c(c0)),lambdaval=0  )[,-c(2:6)])->BN
colnames(BN)<-nombracoeffsmatrix(LM = LD); round(BN,4); 
```

### Constrained least squares all orthants

Here we perform the expensive search over all the space composed of $2^`r nrow(LD)`$ orthants.

```{r}
matrix(nrow=1,lassocono(XM=DATX, YM=DATY, AA=AA, lambdaval=0  )[,-c(2:6)])->BA
colnames(BA)<-nombracoeffsmatrix(LM = LD); round(BA,4); 
```

From the earlier computations, this example is an instance 
where the initial least squares' orthant or neighboring orthants are 
not neccesarily a good starting point as it is evident when compared against the full search.

## Constrained lasso

We now build lasso paths for the cases of neighbor orthants and full orthant. 


### Searching neighbors of least squares orthant

We check the last result using the less expensive search over neighboring orthants of the ordinary least squares orthant (`r c0`). 

```{r}
round(lassocono(XM=DATX, YM=DATY, AA=AA, GG=vecinos(c(c0)), 
                nlambda=7,lmax=16   )[,-c(2:6)],5)->LNE
colnames(LNE)<-nombracoeffsmatrix(LM = LD);  LNE
```

### Searching all orthants

This is the same call of `lassocono` as above except that when the argument `GG` is not given, the search is 
over  the space composed of $2^`r nrow(LD)`$ orthants.

```{r}
round(lassocono(XM=DATX, YM=DATY, AA=AA,
                 nlambda=7, lmax=22  )[,-c(2:6)],5)->LAL
colnames(LAL)<-nombracoeffsmatrix(LM = LD);  LAL
```

\pagebreak

## Plotting the results

The custom built function `graficalo` is used to plot the paths for the two cases, only neighboring orthants and all orthants. To improve the path plotted, we use more values of $\lambda$ than earlier. These are computed with function `seqq`. 

```{r,fig.height=3,fig.width=7}
par(mar=c(4,4,1,1),mfrow=c(1,2))
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY, AA=AA, GG=vecinos(c(c0)), 
                  lambdaval=seqq(maximo=16,cuantos=30)   )[,-c(2:6)], LM=LD,
                  titulo="H neighboring orthants")
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY, AA=AA, 
                lambdaval=seqq(maximo=16,cuantos=30)   )[,-c(2:6)], LM=LD,
                 titulo="H all orthants")
```

We can rescale the restriction matrix by multiplying its positive entries
by a factor. Using a factor bigger than one enlarges the search region and we would 
expect less differences between searching in neighboring orthants against searching over all orthants. This is done below for the factor equal to two and to ten.

```{r,fig.height=3,fig.width=7}
AA2<-rescale(MAT=AA,factor=2); AA10<-rescale(MAT=AA,factor=10); 
par(mar=c(4,4,1,1),mfrow=c(1,2))
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY,  AA=AA2, GG=vecinos(c(c0)), 
          lambdaval=seqq(maximo=24,cuantos=30)   )[,-c(2:6)], LM=LD,
          titulo="H neighboring orthants")
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY,   AA=AA2, 
           lambdaval=seqq(maximo=24,cuantos=30)   )[,-c(2:6)], LM=LD,
           titulo="H all orthants")
```

```{r,fig.height=3,fig.width=7}
par(mar=c(4,4,1,1),mfrow=c(1,2))
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY, AA=AA10, GG=vecinos(c(c0)), 
          lambdaval=seqq(maximo=36,cuantos=30)   )[,-c(2:6)], LM=LD,  
          titulo="H neighboring orthants")
graficalo(Coeffs=lassocono(XM=DATX, YM=DATY, AA=AA10, 
          lambdaval=seqq(maximo=36,cuantos=30)   )[,-c(2:6)], LM=LD,  
           titulo="H all orthants")
```

In this latter case with factor equal to ten the differences between trajectories in the 
two cases become negligible. \pagebreak


Below we exhibit the constraint matrix for this latter case with factor equal to ten, 
```{r}
AA10
```

Here are the Lasso paths for this last scaling factor of ten and the two 
 cases (neighboring orthants and all orthants). The paths are largely similar although
 they are not equal.

```{r}
lassocono(XM=DATX, YM=DATY, AA=AA10, GG=vecinos(c(c0)),
                lambdaval=seqq(maximo=36,cuantos=4)   )[,-c(2:6)]->BBN
colnames(BBN)<-nombracoeffsmatrix(LM = LD);  
lassocono(XM=DATX, YM=DATY, AA=AA10, 
                lambdaval=seqq(maximo=36,cuantos=4)   )[,-c(2:6)]->BBA
colnames(BBA)<-nombracoeffsmatrix(LM = LD);  
round(BBN,5) ## Neighboring orthants
round(BBA,5) ## ALL orthants
```

\pagebreak

# Weak hierarchical Lasso (W)

Here is the constraint matrix for the same model and weak hierarchy. This matrix is obtained with the function `restrict` and setting `type=2`. In this case, the flag `weight=TRUE` 
gives matrix with non zero entries being $\pm 1$. We also give the scaled version of it.

```{r}
restrict(LM=LD,type=2,weight=TRUE)->AA; colnames(AA)<-nombres(LM=LD); AA
rescale(MAT=restrict(LM=LD,type=2,weight=TRUE),factor=10)->AA10; colnames(AA10)<-nombres(LM=LD); AA10
```

The following two paths are for neighboring orthants and for computation using all orthants.
The analyses coincide.


```{r}
lassocono(XM=DATX, YM=DATY, GG=vecinos(c(c0)), AA=AA, 
                lambdaval=seqq(maximo=38,cuantos=4)   )[,-c(2:6)]->BBN
lassocono(XM=DATX, YM=DATY,  AA=AA, 
                lambdaval=seqq(maximo=38,cuantos=4)   )[,-c(2:6)]->BBA
colnames(BBN)<-nombracoeffsmatrix(LM = LD); colnames(BBA)<-nombracoeffsmatrix(LM = LD);
round(BBN,5) ## neighbor orthant
round(BBA,5) ## all orthants
```

\pagebreak

# Strong hierarchical Lasso (S)

This is achieved with the constraint matrix built with `restrict` and `type=1`. The flag
`weight=TRUE` is used as in the weak hierarchy case.


```{r}
restrict(LM=LD,type=1,weight=TRUE)->AA; colnames(AA)<-nombres(LM=LD); AA
rescale(MAT=restrict(LM=LD,type=1,weight=TRUE),factor=100)->AA10; colnames(AA10)<-nombres(LM=LD); AA10
```

Below are two paths built using the matrix `AA` and the two searches over "neighbor orthants" and over "all orthants". It is clear in the plots the suboptimality of neighbor search.

```{r}
lassocono(XM=DATX, YM=DATY, GG=vecinos(c(c0)), AA=AA, 
                lambdaval=seqq(maximo=21.5,cuantos=20)   )[,-c(2:6)]->BBN
lassocono(XM=DATX, YM=DATY,  AA=AA, 
                lambdaval=seqq(maximo=21.5,cuantos=20)   )[,-c(2:6)]->BBA
par(mar=c(4,4,1,1),mfrow=c(1,2)) 
graficalo(Coeffs=BBN,titulo="Neighbor orthant",LM=LD)
graficalo(Coeffs=BBA,titulo="Neighbor orthant",LM=LD)
```

# Appendix 

## Data format

Assume that there are $n$ observations in which a response variable was collected and there are measurements of $d$ covariates available for every value of the response variable.
The data for this analysis consists of three objects.

The **response** vector $Y$ is a column
vector of $n$ rows; there is a **design-model matrix** $X$ of $n$ rows and with $p$ columns.
The columns of $X$ are computed by evaluating polynomial terms at each of covariate values.
These $p$ columns of $X$ correspond to the terms in the polynomial **model** $LM$. 
This model is allocated in a matrix of $p$ rows and with $d$ columns, where $d$ is the number of covariates, and the entries of this matrix are integer numbers which are the exponents that create the polynomial model terms. 


## Main functions from the file `codigofuente.r`

The file has many different utilities, some for creating and handling  polynomial models,
other functions are for handling the model terms, other for plotting analyses. Here I
describe the main functions used for analysis.


## `minconoineq` 

This is the main function, based upon the function `solve.QP` from library `quadprog`. 
For a given model data matrix `XM` with response vector `YM`, and choice of
regularization parameter `lamm` (i.e. $\lambda$), 
the function `minconoineq` 
 performs a quadratic minimization over the orthant given by  diagonal of matrix `CM`.
The objective function is that orthant part of 
standard lasso criterion $$L=||Y-X\beta||_2^2+\lambda||\beta||_1,$$ subject
to constraints in $\beta$ imposed by the constraint matrix `AA` and to the orthant
of `CM`.
The entries of `CM` that identify orthant may be not just $\pm 1$ but could be zero
as well, hence the function has steps that prevent the matrices involved having
zero rows/columns.

The function retrieves the coefficient $\hat\beta^L(\lambda)$ over the specified orthant,
together with the minimal objective value achieved.

## `minconoineqquadr`

This function is a wrapper that runs `minconoineq` over several quadrants specified
by the matrix `GG`. The other entries are `XM`, `YM`, `lamm` and `AA`. If `GG` is not
specified, this function performs a computation over all possible orthants which are
$2^{|LM|}$ (recall that columns of `XM` correspond to terms in $LM$).

If the flag `minimize` is set to `TRUE`, then the function retrieves the best 
result of `minconoineq` over the specified set of orthants `GG`. By default,
`minimize=FALSE` and the function retrieves all the results, one for each orthant.

## `lassocono`

This is a wrapper that runs `minconoineqquadr` (and consequently  `minconoineq`) over 
a choice of values of $\lambda$
specified by one of two ways. The user can give a list of specific
values in `lambdaval`; alternatively the user specifies `lmin`, `lmax` and `nlambda`
to construct a uniformly spaced sequence of values to use. Other entries
`XM`, `YM`, `AA` and `GG` are as per inputs of `minconoineqquadr`. 

The output is the collection of coefficients of the model that minimizes over choice
of orthants `GG`. The output is slightly verbose as for every $\lambda$, it includes
the value of $\lambda$, which orthant was the minimizer, the estimate 
$\hat\beta^L(\lambda)$ and the minimal objective value achieved.

## `hassediagram`

This function creates the simplified Hasse diagram. Its input is `LM`,
the matrix in which each row is a term and entries are exponents of the
model.

Each row of the output corresponds to a term (row) in `LM`. In each row,
nonzero numbers are pointers (indexes) which other model terms depend on 
the current row, via simple product of unit exponents and thus the number
of columns equals the number of variables (in turn, number of columns in `LM`).

## `restrict`

This function generates the matrix of restrictions to be used as input `AA` 
in the above functions. The following are inputs for this function, `LM` which is the
specification of polynomial terms in the model, and `type` that determines
the hierarchy to be used. If `type=1` strong hierarchy (S) adding over descendants
is done, if `type=2` then return weak hierarchy (W) adding over parents and
otherwise `type=3` return general strong hierarchy (H). The other
input is the flag `weight`. By default `weight=TRUE` and all non-zero entries 
are $\pm 1$. This flag does not matter for `type=3`.
This function calls automatically `hassediagram` so no need to create a
Hasse diagram separately.

The output is a matrix with the weights for the inequalities. It has as many
columns as the terms (rows) in `LM` and as many rows as the type of hierarchy.
If (S,  `type=1`) the number of rows is the number of nodes with are parents;
for (W,  `type=2`) it is the number of nodes which are descendants, and
for (H,  `type=3`) it is the number of edges in the Hasse diagram.

## `rescale`

This function is to scale the positive entries of matrix `MAT`. Depending on 
the value of input `factor`, it is possible
to make the search region less constrained for `factor` bigger than one and
more constrained for such value less than one. Usually we want to make the analysis
less constrained. A logical flag `override` controls whether this output will be forced
to be exactly equal to `factor`, otherwise it is the product of the maximum positive
entry by `factor`.

The output is the scaled matrix `MAT` that will be used 
as matrix of restrictions `AA`.

## `vecinos`

This function takes as input `CC` a vector of entries taking values of $\{0,\pm 1\}$.

The output is a matrix output with as many columns as entries in `CC` and the
rows are sign changes of `CC` one at a time, including the
original `CC`. This is to be used in a simple, low cost orthant search. This
is the method refered to as the 'neighbor orthant' method.


## `graficalo`

This is a call to plot the lasso path, stemming from the output of `lassocono`. 
Its main input
is `Coeffs` which by default (controlled by flag `extras`) is assumed to have $\lambda$ in 
the first column then columns for the
coefficients $\hat\beta^L(\lambda)$ and in the final column the criterion $L$.
If the table for `Coeffs` only has the coefficients $\hat\beta^L(\lambda)$ 
available, the flag `extras` should be set to `FALSE`.

By default the path are plotted against percentage of shrinkage $s=s(\lambda)$,
but they could be plotted against $\lambda$ by setting the flag `plotlambda=TRUE`.
The flag `simplito` by default allows for very simple axis labels but could
create more complex labels if desired by switching to `simplito=TRUE`.
If model `LM` is given, the plot adds labels to each path. 
